---
title: 'Regression Models Project: Motor Trend MPG Analysis'
author: "Telvis Calhoun"
date: "February 26, 2016"
output:
  pdf_document: default
  html_document:
    keep_md: yes
---

## Overview

In this project, we explore the `Motor Trend Car Road Tests` (mtcars) dataset. We'll analyze this dataset to answer the following questions. 

1. “Is an automatic or manual transmission better for MPG”
2. "Quantify the MPG difference between automatic and manual transmissions"

## Exploratory Analysis

First, lets load libraries and datasets used in the analysis.

```{r, message=FALSE}
library(datasets)
library(ggplot2)
library(dplyr)
data("mtcars")
```

A quick summary of the data shows `mtcars` dataset 11 variables. For this analysis, we will investigate the Miles/(US) gallon `mpg` as a function of the Transmission type `am`.

```{r echo=TRUE,cache=TRUE}
summary(mtcars)
```

Let's create a factor variable called `am_factor` that will show the strings 'automatic' where `am == 1` and 'manual' where `am == 0`.

```{r echo=TRUE,cache=TRUE}
mtcars <- mutate(mtcars, am=factor(ifelse(am==1, 'automatic', 'manual')))
table(mtcars$am)
```

## Model Selection

Before we can characterize the relationship between `mpg` and the transmission type (`am`), we must first search for other variables that can distort, or confound the relationship between the variables. First we will "get our hands dirty" and quantify the percentage change in the `am` coefficient when we adjust for all other variables.

```{r echo=TRUE,cache=TRUE}
# baseline
a <- summary(lm(mpg ~ am, data=mtcars))$coef[2]

# calc percent change in the baseline for each covariate
rbind(
  c('baseline', 'cyl', 'disp', 'hp', 'drat', 'wt', 'qsec', 'vs', 'gear', 'carb'),
  round(c((a-a)/a,
    100 * (a - summary(lm(mpg ~ am + cyl, data=mtcars))$coef[2])/a,
    100 * (a - summary(lm(mpg ~ am + disp, data=mtcars))$coef[2])/a,
    100 * (a - summary(lm(mpg ~ am + hp, data=mtcars))$coef[2])/a,
    100 * (a - summary(lm(mpg ~ am + drat, data=mtcars))$coef[2])/a,
    100 * (a - summary(lm(mpg ~ am + wt, data=mtcars))$coef[2])/a,
    100 * (a - summary(lm(mpg ~ am + qsec, data=mtcars))$coef[2])/a,
    100 * (a - summary(lm(mpg ~ am + vs, data=mtcars))$coef[2])/a,
    100 * (a - summary(lm(mpg ~ am + gear, data=mtcars))$coef[2])/a,
    100 * (a - summary(lm(mpg ~ am + carb, data=mtcars))$coef[2])/a
  ), digits=2)
)
```

The output shows that `cyl`, `disp`, `hp`, `drat`, `wt` all change the `mpg` coeffient by greater than +/- 25%. Now let's generate nested models for these `5` variables using and evaluate them using a nested likelihood ratio tests.

```{r echo=TRUE,cache=TRUE}
fit1 <- lm(mpg ~ am, data=mtcars)
fit2 <- lm(mpg ~ am + cyl, data=mtcars)
fit3 <- lm(mpg ~ am + cyl + disp, data=mtcars)
fit4 <- lm(mpg ~ am + cyl + disp + hp, data=mtcars)
fit5 <- lm(mpg ~ am + cyl + disp + hp + drat, data=mtcars)
fit6 <- lm(mpg ~ am + cyl + disp + hp + drat + wt, data=mtcars)
anova(fit1, fit2, fit3, fit4, fit5, fit6)
```

The results show that models 2, 4 and 6 have the greates significance. This suggests that `disp` and `drat` should be excluded from the model. The comparison below that good model lowest p-value with the variables: `am`, `cyl`, `hp` and `wt`.

```{r echo=TRUE,cache=TRUE}
fit1 <- lm(mpg ~ am, data=mtcars)
fit7 <- lm(mpg ~ am + cyl + hp + wt, data=mtcars)
```

The `r.squared` value for this model indicates that 83% of the total variablity is explained by the linear relationship between the `mpg` and `am`, `cyl`, `hp` and `wt`.

```{r echo=TRUE,cache=TRUE}
summary(fit7)$adj.r.squared
```

Finally, we plot the residuals to search for a pattern in the residuals vs the fitted (yhat) values. The plot shows a slight curve in the values - but nothing too bad.

```{r echo=FALSE,cache=TRUE}
fit_lm <- lm(mpg ~ am + cyl + hp + wt, data=mtcars)
df <- data.frame(e=resid(fit_lm), yhat = predict(fit_lm))
g <- ggplot(df, aes(yhat, e), title="foo") + 
  geom_point() + 
  geom_smooth(method="loess") +
  ggtitle("Residuals Vs. Fitted (linear model)") +
  xlab("Fitted Miles per Gallon (mpg)") +
  ylab("Residuals (mpg)")

print(g)
```

## Automatic Vs. Manual Comparison

Now that we have a model containing the necessary covariates, let's calculate the change in `mpg` for transmission type after adjusting for Number of Cylinders (`cyl`), Gross Horsepower (`hp`) and weight (`wt`). The results show that the `mpg` for a manual transmission decreases by `1.48 mpg` holding other variables constant.

```{r echo=TRUE,cache=TRUE}
summary(fit_lm)$coef
```

However, the confidence interval for `am` ranges from `-4.4` to `1.47`. Because it includes 0, we cannot reject the null hypothesis after  adjusting for Number of Cylinders (`cyl`), Gross Horsepower (`hp`) and weight (`wt`).

```{r echo=TRUE,cache=TRUE}
confint(fit_lm)
```


## Conclusion
We show that we most accurately model the relationship between `MPG` and `transmission type` using a `poisson` generalized linear model. This model has lower residual error than `linear` model. `TODO: FIXME` : The results show that an automatic transmission has `0.0%` greater fuel efficiency than manual transmission. However, the results show that the difference in fuel efficiency decreases by `0.0%` when we adjust for the number of cylinders.
